Effects

### 2020-01-08 doc/Effects.Rmd
Wrapping up Effects Report, fixing some bugs

I am testing the quality effects, since it could be a source of error and potentially biased the predicted gwRR if some mice had bad technical noise.

@@

```{r MLH1.histogram.2, echo=FALSE, include=FALSE}
#thse plots not shown
#remake the pooled MLH1 histgoram plot from above but facet by quality

@@

These plots show that generally there is a positive relationship between good quality (1) and more MLH1 counts.

I should focus on addressing if there are differences in the relationship of quality-MLH1 counts across sexes.

@@

Examine image quality or other scores to test how well the python scripts segment the SC area for all the images.

```{r SC.qual, echo=FALSE}
#use the SC table

@@

```{r SC.hitograms, echo=FALSE, fig.width = 4, fig.height=1}
#try adjusting the size for thi..

@@

```{r more.SC.hist, echo=FALSE}

@@

```{r sc.qual.plots, echo=FALSE}
skel.mm <- mouse.table.mean.skel.obs_qual[mouse.table.mean.skel.obs_qual$sex == "male",]
skel.mm <- skel.mm[skel.mm$mean.skel < 50000,]

@@

```{r G.plot, echo=FALSE}
Bin.qual.effect.plot.mm.G <- ggplot(skel.mm[skel.mm$strain == "G",],
                           aes(y=mean.bin, x=quality, size= ncells, color=strain))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Quality and mean SC (bin) ")+facet_wrap(~mouse)

@@

The mouse 22jun15_G_m2 has strange pattern, investigate these outputs

@@

```{r female.skel.plot, echo=FALSE}

@@

skel.qual.effect.plot.ff <- ggplot(skel.ff, aes(y=mean.bin, x=quality, color=(strain) ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Male quality and mean CO")+facet_wrap(~subsp)

@@

skel.qual.effect.plot.ff.MSM <- ggplot(skel.ff[skel.ff$strain == "MSM",], aes(y=mean.bin, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Male quality and mean CO")+facet_wrap(~mouse)

skel.qual.effect.plot.ff.G <- ggplot(skel.ff[skel.ff$strain == "G",], aes(y=mean.bin, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Male quality and mean CO")+facet_wrap(~mouse)

skel.qual.effect.plot.ff.PWD <- ggplot(skel.ff[skel.ff$strain == "PWD",], aes(y=mean.skel, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Male quality and mean CO")+facet_wrap(~mouse)

@@

The above plots are good for detecting outliers

@@

Some papers have documented age effects for male genome-wide recombination rate [@zelazowski2017age].

The initial analysis uses plots. This should be double checked with a general linear model as well.

@@

I made 3 models for 3 groupings of mice, dom, musc.high and musc.low

$$mouse .mean CO ~=~ age + strain + \varepsilon$$

@@

ADD euthdate?

Dom, no significant effects for age.

Musc low, no significant effect for strain or age.

Musc high, MSM and SKIVE have significant age effects on mean CO count

SKIVE has the greatest spread in ages for single mice (7,16,41 weeks)
There are more SKIVE males which I haven't quantified (at least 3, which are in a good age range)

@@

I like this plot and find it effective. Can I do a similar set up for other factors (quality, ncells?)

@@

Compare embryonic to neonate (logistic regression). I don't think my embyonic stage estimates are very accurate, so I'll bin all embryos into 1.
Note all strains have both embryo and neonate samples. Subset for just, WSB, G, LEW, and PWD.

@@

The mouse curves are more slanted, but the logistic regressions are not significant. The lg for WSB mouse averages isn't significant.

@@

Check if there is a maternal age effect for females and males.
clea.Meta.Data doesn't have maternal age data.

Collect data, get a decent set / large set of mice with maternal ages (mostly want to look at females).

@@

mat.age_MLH1_mouse

@@

there should be plots above this
Based on the histograms of ages and the range across categories isn't very wide -- so not sure what the best test is.

@@

When this project was started the main mouse room for vendor purchased mice was Biotech. The mouse room changed in summer 2017. The Gough mice were always housed in Charmony. Of the final dataset there are three strains had mice housed in different rooms; WSB, LEW, and MSM.

The main questions to address in this section: is there a significant GxE effect due to room?  do the different rooms added noise / bias the estimates per strain

A room effect might cause all of the strains to be biased in the same direction.

```{r ROOM.effect.setup, echo=FALSE}
#pull data from the 3 strains
room.effect.DF <- MLH1_data[(MLH1_data$strain == "WSB") | (MLH1_data$strain == "LEW") | (MLH1_data$strain == "MSM"), ]

Looking at my records, there are limited comparisons (not all strains were raised in all 3 rooms).

WSB female - Biotech vs Charmony

LEW female and male, MSM female and male - Biotech and MSC

look at boxplots first

```{r room.effect.tests, echo=FALSE}
library(ggsignif)

room.boxplots.cell <- ggplot(data=FULL.room.data, aes(x = room, y = nMLH1.foci))+geom_jitter()+geom_boxplot(alpha=.5)+facet_wrap(~category)+
   #test bars
  ggsignif::geom_signif(
    data=geom.sig_df.cells, 
    aes(xmin=group1, xmax=group2, annotations=p.adj, y_position=y_pos),
    manual=TRUE)+
  ggtitle("Boxplots of room effects,\nsingle cell level, adj p value")
room.boxplots.cell

room.boxplots.mouse <- ggplot(mouse.level.room.data, aes(x=room, y=mean.MLH1))+geom_boxplot(alpha=.5)+facet_wrap(~category)+geom_jitter()+
  #test bars
geom_signif(
    data=geom.sig_df.mouse, 
    aes(xmin=group1, xmax=group2, annotations= p.adj, y_position= y_pos ), manual=TRUE) +  ggtitle("Boxplots of room effects,\nmouse level, adj p value")
room.boxplots.mouse

#MORE TESTS TO consider
#the boxplots show different ranges
#room.boxplots.OG <- ggplot(mergy.room.data.OG)
#tukey / post hoc paired comparison
```
the brackets in the plots above reflect the output of 'compare_means()' The default is wilcox.test, I've been swtiching back between showin raw and adjusted p values. (look up the warning about the cols parameter.)

The values look a little strange to me, I don't understand the p.adj values yet, (pvalues muliutpplied by the number of comparisons). To me it seems like there are 5 t.tests, . I don't know why there are only 3 values of adjusted p-values)

# Litter effect

@@

# Dissection spread Effect

intro for why I care about dissection.
right nnow the only independant variable is month dissected. Should I enter infor about PFA? time in humid chamber?

```{r dissection.effect.data.setup, echo=FALSE}

@@

```{r plot.dissection, echo=FALSE}

@@

```{r dissection.female, echo=FALSE}

@@

## Total SC

Testing the effects of mean total SC per mouse with when they were dissected.

follow up to peromyscus findings, Where summer caught mice had lower SC area per cell, compared to winter and fall caught. But there were few data points.

```{r total.sc, echo=FALSE}
#check out effects of month dissected on SC
#load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_12.20.19.RData")#fixing bugs, formating
#from total SC

@@

So, there doesn't seem to be an effect of month dissected on SC area across mice. This might be the start of testing for 'technical effects / humidity effects'. BUT, I don't really have month by month dissections of mice of the same age, ect.

> *month dissected*

@@

# Stain Effects

Intro for staining effects. at least 3 slides were made per mouse, giving a potential of 3 different stains. The staining batch might have effects on the background noise and brightness of the MLH1 foci.

collect data on mice with mulitple spreads (sp1, sp2 -- make ddply table that counts a gerp for 'sp1', sp2, ect)

> *staining batches*

Compare observations from the same mouse, but different MLH1 stains.  First need to identify data that can be used.

```{r stain.effects.SC, eval=FALSE}
#/src/TotalSC.R
#compile this dataset

>>>>

> Main points

1. There is a positive relationship with quality and MLH1 count (better cells have more MLH1)

2. The effect of Quality on mean MLH1 count is larger in males

> Open quesitons

 - Are quality score less accurate/consistant for females?

 - Should I weight mouse COs averages by the quality ratio?

 - Exclude lowest quality cells in males, but keep them in female dataset?

 - Does the effect for males still hold when Q5 cells are removed?

@@

```{r MLH1.histogram.2, echo=FALSE, eval=FALSE,include=FALSE}
#thse plots not shown
#remake the pooled MLH1 histgoram plot from above but facet by quality

@@

These plots show that generally there is a positive relationship between good quality and more MLH1 counts.

```{r qual.effect.sex, echo=FALSE}
#compare the quality effect by sex
#an issue with comparing sexes is they have different ranges of MLH1 means
#do other analysis

#plots show lm for observations from mouse.table.mean.MLH1.obs_qual
#glm

blah.lm.ff <- lm( mean.MLH1 ~ quality + ncells + subsp + strain, data=ff)
#quality bin is significant to predict count in ff (remeber these are binned averages)
# -0.85303,  3.61e-14 ***

blah.lm.mm <- lm( mean.MLH1 ~ quality + ncells + subsp + strain, data=mm)
#quality co Eff,   -8.863e-01, < 2e-16 ***

mm.low <- mm[!( (mm$strain == "MSM") | (mm$strain == "PWD") ), ]

blah.lm.mm.low <- lm( mean.MLH1 ~ quality + ncells + subsp + strain, data=mm.low)
#quality sig, so strain (unlike female (probably bc greater range))

# -8.881e-01,  < 2e-16 ***

#the coefficient is 10x higher in males, p value max-out low 
#females the quality co-E is much smaller and p.value larger  
#quality effect on mean co's is larger in males

I ran lm of the data from the plots above and found that the coefficient is 10x higher in males, p value max-out low and in females the quality co-efficicent is much smaller and p.value larger.

@@

In progress. Examine image quality or other scores to test how well the python scripts segment the SC area for all the images.

```{r SC.qual, echo=FALSE, eval=FALSE}
#use the SC table

@@

```{r SC.hitograms, echo=FALSE, eval=FALSE, fig.width = 4, fig.height=1}
#try adjusting the size for thi..

@@

```{r more.SC.hist, echo=FALSE, eval=FALSE}

@@

```{r sc.qual.plots, echo=FALSE, eval=FALSE}
skel.mm <- mouse.table.mean.skel.obs_qual[mouse.table.mean.skel.obs_qual$sex == "male",]
skel.mm <- skel.mm[skel.mm$mean.skel < 50000,]

@@

```{r G.plot, echo=FALSE, eval=FALSE}
Bin.qual.effect.plot.mm.G <- ggplot(skel.mm[skel.mm$strain == "G",],
                           aes(y=mean.bin, x=quality, size= ncells, color=strain))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Quality and mean SC (bin) ")+facet_wrap(~mouse)

@@

The mouse 22jun15_G_m2 has strange pattern, investigate the Python SC outputs.

@@

```{r female.skel.plot, echo=FALSE, eval=FALSE}

@@

skel.qual.effect.plot.ff <- ggplot(skel.ff, aes(y=mean.bin, x=quality, color=(strain) ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Female quality and mean CO")+facet_wrap(~subsp)

@@

skel.qual.effect.plot.ff.MSM <- ggplot(skel.ff[skel.ff$strain == "MSM",], aes(y=mean.bin, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Female quality and mean CO")+facet_wrap(~mouse)

skel.qual.effect.plot.ff.G <- ggplot(skel.ff[skel.ff$strain == "G",], aes(y=mean.bin, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Female quality and mean CO")+facet_wrap(~mouse)

skel.qual.effect.plot.ff.PWD <- ggplot(skel.ff[skel.ff$strain == "PWD",], aes(y=mean.skel, x=quality, color=(strain), size=ncells ))+geom_point()+ 
  scale_color_manual(values=colors_of_strains)+ #this list in setupdata
  geom_smooth(method='lm',alpha=0.1, size=0, span=0.5 )+ggtitle("Female quality and mean CO")+facet_wrap(~mouse)

@@

@@

Some papers have documented age effects for male genome-wide recombination rate [@zelazowski2017age] (after maturity, there's a negative correlation with age and recombination).

> Main Points

 1. Weak evidence for age effect on MLH1 counts for male mice.

 - Dom, no significant effects for age.

 - Musc low, no significant effect for strain or age.

 - Musc high, no significant age effects on mean CO count

My previous dataset showed age effect for SKIVE males due to unbalanced data. There were only 3 mice of 7,16, and 41 weeks. The models using this dataset had signifcant age effects. The updated data set now has ~4 mice of 7 weeks and no significant age effect.

I tested the age effects with 3 sets of models,  for 3 groupings of mice: dom, musc.high and musc.low

@@

$$mouse .mean CO \sim age + strain + \varepsilon$$

@@

The plot below shows mouse averages as single points, showing the mean MLH1 and age in weeks. The lines and shading show the smoothed lm for each strain.

@@

The models and mostly flat slopes indecate that there isn't a strong age effect on MLH1 for these mice. (there might be a KAZ male outlier (10jul19_KAZ_m3), double check this). Change the colors for KAZ, TOM, AST, and CZECH they're too similar.

@@

Female ages most logically group into neonate or embryo, so I choose to apply a logistic regression to test for age effects. Not all strains have both embryo and neonate female samples, so just, WSB, G, LEW, and PWD are examined.

> Main Point

 - MLH1 counts don't differ between neonates and embryos

Below are the tests for the model (can CO number predict the age stage). I ran these comparisons for cell count and mouse averages. The plots show the points and smoothed curves with age on the y axis, 0 = embryo and 1 = neonate.

$$ Age .Stage \sim CO .number$$

@@

The mouse curves are more slanted, but the logistic regressions are not significant. None of the indiviual strains had significant p values when compared by themselves.

When this project was started the main mouse room for vendor purchased mice was Biotech. The mouse room changed in summer 2017. The Gough mice were always housed in Charmony. Of the final data set there are three strains had mice housed in different rooms; WSB, LEW, and MSM. (not balanced/ limited comparisons / )

The main questions to address in this section: is there a significant GxE effect due to room?

> Main Points

 - There is some evidence for room effect on MLH1 counts per cell for WSB and LEW female and LEW and MSM male.

 - No support of room effect for mouse means.

For the cell counts of MLH1, Two of three are in the same direction. Sample size might be having an effect.

There might be more mice/data that can be added to these tests, need to comfirm the move date and verify with lab note book.

```{r ROOM.effect.setup, echo=FALSE, warning=FALSE}
#pull data from the 3 strains
room.effect.DF <- MLH1_data[(MLH1_data$strain == "WSB") | (MLH1_data$strain == "LEW") | (MLH1_data$strain == "MSM"), ]

```{r room.effect.tests, echo=FALSE, warning=FALSE}
library(ggsignif)

mouse.level.room.data <- mouse.level.room.data[!(is.na(mouse.level.room.data$mean.MLH1)|mouse.level.room.data$mean.MLH1==""),]

#note the mouse level tests are all throwing errors because not enough samples

#Wlcx.ROOM_WSB.female  <- wilcox.test(FULL.room.data$mean.MLH1[ ( (FULL.room.data$category == "WSB female") & (FULL.room.data$room == "B2420") ) ], 
#            FULL.room.data$mean.MLH1[ ( (FULL.room.data$category == "WSB female") & (FULL.room.data$room == "Charmony") ) ]  ) #raws for compare means match

Wlcx.ROOM_LEW.male  <- wilcox.test(FULL.room.data$nMLH1.foci[ ( (FULL.room.data$category == "LEW male") & (FULL.room.data$room == "B2420") ) ], 
           FULL.room.data$nMLH1.foci[ ( (FULL.room.data$category == "LEW male") & (FULL.room.data$room == "MSC 3040") ) ]  )

room.boxplots.cell <- ggplot(data=FULL.room.data, aes(x = room, y = nMLH1.foci))+geom_jitter()+geom_boxplot(alpha=.5)+facet_wrap(~category)+
   #test bars
  ggsignif::geom_signif(
    data=geom.sig_df.cells, 
    aes(xmin=group1, xmax=group2, annotations=p, y_position=y_pos),
    manual=TRUE)+
  ggtitle("Boxplots of room effects,\nsingle cell counts, raw p value")
room.boxplots.cell

room.boxplots.mouse <- ggplot(mouse.level.room.data, aes(x=room, y=mean.MLH1))+geom_boxplot(alpha=.5)+facet_wrap(~category)+geom_jitter()+
  #test bars
geom_signif(
    data=geom.sig_df.mouse, 
    aes(xmin=group1, xmax=group2, annotations= p, y_position= y_pos ), manual=TRUE) +  ggtitle("Boxplots of room effects,\nmouse means, raw p values")
room.boxplots.mouse

#MORE TESTS TO consider
#the boxplots show different ranges
#room.boxplots.OG <- ggplot(mergy.room.data.OG)
#tukey / post hoc paired comparison
```

The brackets in the plots above reflect the output of 'compare_means()' which uses a wilcox.test.

I double checked the wilcox p values for wilcox.test and compare_means, they're similar.

WSB single wilcox test pvalue is  Wlcx.ROOM_WSB.female$p.value` which is the same as the raw from compare_means(). The compare_means table shows that WSB is multiplied by 2.

The values look a little strange to me, I don't understand the p.adj values yet, (p values multiplied by the number of comparisons). To me it seems like there are 5 t.tests, . I don't know why there are only 3 values of adjusted p-values)

@@

> Main Point

 - Qualitatively, there is unlikely to be a maternal age effect, but these data are underpowered.

@@

@@

@@

For the topics below, Maternal age, litter and dissection effect are all related, I need to outline/distinguish how and decide if these are worth further investigation.

## Litter effect

@@

## Dissection spread Effect

right now the only independent variable is month dissected. Should I enter info about PFA? time in humid chamber?

```{r dissection.effect.data.setup, echo=FALSE, warning=FALSE}

@@

```{r plot.dissection, echo=FALSE, warning=FALSE}

@@

```{r dissection.female, echo=FALSE, warning=FALSE}

@@

### Total SC

Testing the effects of mean total SC per mouse with when they were dissected. This is mean to be a follow up to peromyscus findings, Where summer caught mice had lower SC area per cell, compared to winter and fall caught. But there were few data points.

```{r total.sc, echo=FALSE, warning=FALSE}
#check out effects of month dissected on SC
#load(file="C:/Users/alpeterson7/Documents/MLH1repo/data/MLH1/MLH1_data_setup_12.20.19.RData")#fixing bugs, formating
#from total SC

@@

So, there doesn't seem to be an effect of month dissected on SC area across mice. This might be the start of testing for 'technical effects / humidity effects'. BUT, I don't really have month by month dissections of mice of the same age, etc.

@@

# Stain Effect

At least 3 slides were made per mouse and I rarely included slides from the same mouse when running the statining protocol. Different staining batch might have effects on the background noise and brightness of the MLH1 foci (these could also be affected by post processing of images to a lesser extent). A siginificant stain effect would indecate technical error / I wasn't consistant in running the staining protocol).

**collect data set for mice with multiple spreads (sp1, sp2 make ddply table that counts a gerp for 'sp1', sp2, ect)**

These data might also be underpowered, since I mostly used 1 spread and it was rare for me to collect over ~20 cells from multiple spreads from the same mouse.
Compare observations from the same mouse, but different MLH1 stains.  First need to identify data that can be used.

```{r stain.effects.SC, eval=FALSE}

#compile this dataset
